A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-01 paper.tex
Fill out discussion a bit

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but, crucially,
where each individual is represented by two nodes (for diploids).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

@@

% % JK: This is very rough, just jotting down the basic ideas,
% % will need substantial revision.
% It is not clear that such complete information will always
% (or indeed ever) be available from inferences~\cite{shipilina2022origin}.
%  We may be
% able to precisely identify the location and timing
% of recent recombinations (particularly if detailed pedigree
% information is available) but beyond this, the information
% required simply isn't present. When sampling from a
% process such as the Sequentially Markov
% Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
% like ARGweaver~\citep{rasmussen2014genome}, it is true that
% we do obtain a fully resolved ARG with precise
% information about recombination, but it is reasonable
% to question how much information from the data is used
% to support older recombinations, in particular.

% % This idea of a continuum of information about recombination
% % being inferrable and representable is consistent with the ideas
% % discussed by~\cite{shipilina2022origin}. There a haplotype block
% % is synonymous with coalescent nodes in the ARG, which we can generalise
% % as nodes in a gARG with at least two children. In principle, these
% % nodes are identifiable by the mutations that occur on the outbound
% % edges of the node.

@@

% \section*{Equivalance of ARGs and the set of local trees}
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is not as simple as it may seem.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.

% Firstly, the internal nodes in the local trees must be labelled
% (i.e.\ marked with the identity of the corresponding ARG node).
% If the nodes are not labelled, we must infer their identity
% from the tree topology, which is both
% expensive to compute and imprecise. For example
% we could infer that node \textsf{k} is shared in the second
% and third trees of in Fig.~\ref{fig-simplification}D
% because it subtends identical subtrees in both,
% but it would not be straightforward to infer that it
% is also the root of the first tree.
% % JK: Does this need to be said?
% % (One may use the time of a node as a proxy for identity,
% % but it is not reliable as we can have multiple nodes at the
% % same time, e.g. with time discretisation.)

% The precision of the information about recombination that
% can be represented in the trees depends the ARG itself.
% Secondly, the local trees must contain ``unary'' nodes.
% [Some discussion].
% With these two conditions met, there is a direct correspondance
% between a gARG and the trees.

@@

The necessity of complete precision about all recombination
events under the eARG encoding has been
conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
For example, \citet{brandt2021evaluation} define a ``full'' ARG as
``a structure that encodes all
coalescence and recombination events''. While there are certainly cases
when inferring such fully precise ARGs is useful, for example when
studying the details of a small number of recombination
events~\citep{rasmussen2022espalier}, it is worth considering
whether sufficient information exists to support such inferences
and indeed whether this precision is required (or indeed useful
for) downstream applications.

[Discuss information about recomb]
\citet{shipilina2022origin}

% % Do downstream applications actually make use of such full ARGs?
Downstream applications using estimated ARGs to date rarely
make use of such complete recombination information.
The \relate\ selection statistic~\citep{speidel2019method}
computes a $p$-value for selection [FIXME] for each
local tree independently by comparing the rates of coalescence
at various time-points. For cases like strong positive selection,
where we expect there to be a clear signal in the local tree,
such tree-by-tree methods can be simple and powerful.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in a ``full'' ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
for ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
[MORE applications]

% A major strength of this approach is that it can be used on
% any gARG, ranging from estimates containing relatively little
% information about recombination (e.g., as estimated by \relate)
% to ``full'' ARGs containing complete recombination information
% (e.g., as estimated by \kwarg), and will fully utilise
% any between-tree correlation information that is present.

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While the coalescent with recombination is a very useful and
robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
substantially \emph{larger} than the usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with large samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination, but which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-Cov-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.

>>>>

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
\cite{shipilina2023origin}; we avoid term haplotype because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] and the existance of a particular 
haplotype over many generations. A ``genome'' in our usage corresponds to the 
genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We consider only patterns of genetic inheritance, with mutational 
processes being entirely separate.

A gARG is then a graph in which nodes represent
monoploid genomes and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but, crucially,
where each individual is represented by two nodes (for diploids).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

\cite{mathieson2020ancestry} have a slightly different definition of an ARG,
which they say ``is a subset of the pedigree... contain[ing] only those edges along which
inherited segments of DNA have been transmitted, and only those nodes
corresponding to ancestors in which there was a recombination or coalescence
event.'' Thus, while their definition is decoupled from the coalescent
process it is defined by events.

@@

% % JK: This is very rough, just jotting down the basic ideas,
% % will need substantial revision.
% It is not clear that such complete information will always
% (or indeed ever) be available from inferences~\cite{shipilina2023origin}.
%  We may be
% able to precisely identify the location and timing
% of recent recombinations (particularly if detailed pedigree
% information is available) but beyond this, the information
% required simply isn't present. When sampling from a
% process such as the Sequentially Markov
% Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
% like ARGweaver~\citep{rasmussen2014genome}, it is true that
% we do obtain a fully resolved ARG with precise
% information about recombination, but it is reasonable
% to question how much information from the data is used
% to support older recombinations, in particular.

% % This idea of a continuum of information about recombination
% % being inferrable and representable is consistent with the ideas
% % discussed by~\cite{shipilina2023origin}. There a haplotype block
% % is synonymous with coalescent nodes in the ARG, which we can generalise
% % as nodes in a gARG with at least two children. In principle, these
% % nodes are identifiable by the mutations that occur on the outbound
% % edges of the node.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is not as simple as it may seem.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.

Firstly, the internal nodes in the local trees must be labelled
(i.e.\ marked with the identity of the corresponding ARG node).
If the nodes are not labelled, we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.
% JK: Does this need to be said?
% (One may use the time of a node as a proxy for identity,
% but it is not reliable as we can have multiple nodes at the
% same time, e.g. with time discretisation.)

The precision of the information about recombination that
can be represented in the trees depends the ARG itself.
Secondly, the local trees must contain ``unary'' nodes.
[Some discussion].
With these two conditions met, there is a direct correspondance
between a gARG and the trees.

\cite{shipilina2023origin} discuss the same ideas, and note that 
``full ARG ... contains more information than the series of tree
sequences along the genome''. They view the ``full ARG'' as an
eARG, and the ``series of tree sequences'' is their 
% jk: probably should delete this. It *is* muddled though
(somewhat muddled) 
term for the sequence of local trees along the genome. We can see
that their statment is true, \emph{if} the local trees contain
no unary nodes. However, as we have shown, if the local trees 
contain labelled unary nodes, then  there is precisely as 
much information in them as an ARG. Whether such unary nodes
are \emph{observable} is an entirely separate question,
of course.

@@

The necessity of complete precision about all recombination
events under the eARG encoding has been
conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
For example, \citet{brandt2021evaluation} define a ``full'' ARG as
``a structure that encodes all
coalescence and recombination events''. While there are certainly cases
when inferring such fully precise ARGs is useful, for example when
studying the details of a small number of recombination
events~\citep{rasmussen2022espalier}, it is worth considering
whether sufficient information exists to support such inferences
and indeed whether this precision is required, or useful,
for downstream applications.

The amount of information about recombination that is inferrable 
is the subject of much research
has a long history [summarise this in a couple of sentences?]
\cite{shipilina2023origin} discuss the idea of a ``haplotype block''
(equivalent to the ``bricked tree sequence'' of
\cite{nowbandegani2023extremely}), where we consider the unique sets of 
samples that coalesce at a node in the ARG over a particular 
genome interval, and the fundamental limitations they place 
on ARG inference.

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully precise ARGs
from data,
we should also consider the value that such precise estimates provide
to downstream applications. 
To date, few downstream applications using estimated ARGs 
make use of such complete recombination information.
The \relate\ selection statistic~\citep{speidel2019method}
computes a $p$-value for selection [FIXME] for each
local tree independently by comparing the rates of coalescence
at various time-points. 
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in a ``full'' ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
for ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
[MORE applications]
The main application for fully precise ARGs thus far has been
to be to compute a likelihood under the coalescent with 
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian}, 
which currently requires the details of all recombination
events to be known.

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
substantially \emph{larger} than the usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with large samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination, but which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-Cov-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.
